---
title: "Data 607 Project 2"
author: "Ajay Arora"
date: "September 18, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#1. Import libraries
```{r}
library(stringr)
library(tidyr)
library(dplyr)
library('scales')
library(ggplot2)
library(kableExtra)

```

## R Markdown

## First dataset summary:

Summary of the Survey Data

We are studying the species repartition and weight of animals caught in plots in our study area. The dataset is stored as a comma separated value (CSV) file. Each row holds information for a single animal, and the columns represent:  

The data was downloaded from the following site: https://ndownloader.figshare.com/files/2292169

References: https://r4ds.had.co.nz/tidy-data.html

            https://www.google.com/search?q=R+gather()+expalined&ie=&oe=

            https://datacarpentry.org
            
            https://uoftcoders.github.io/rcourse/lec04-dplyr.html            

Column 	Description

record_id 	Unique id for the observation

month 	month of observation

day 	day of observation

year 	year of observation

plot_id 	ID of a particular plot

species_id 	2-letter code

sex 	sex of animal ("M", "F")

hindfoot_length 	length of the hindfoot in mm

weight 	weight of the animal in grams

genus 	genus of animal

species 	species of animal

taxon 	e.g. Rodent, Reptile, Bird, Rabbit

plot_type 	type of plot


#dplyr functions:

    select(): subset columns
    filter(): subset rows on conditions
    mutate(): create new columns by using information from other columns
    group_by() and summarize(): create summary statisitcs on grouped data
    arrange(): sort results
    count(): count discrete values

#tidyr functions:
    spread(): opposite of gathering.You use it when an observation is scattered across multiple rows.
    gather(): collects a set of column names and places them into a single "key" column.
    

#Retrieve Data File for Dataset 1.

```{r}
dataset0 <- read.csv(file="https://raw.githubusercontent.com/AjayArora35/Data-607-Project-2/master/combined.csv", header=TRUE)
head(dataset0)

```
#Data Clean -- Get rid of NAs and any blank data
```{r}
dataset1 <- dataset0 %>% filter(!is.na(hindfoot_length)) %>% filter(!is.na(sex)) %>% filter(!is.na(weight)) %>% filter(!(sex == '')) 

```
#Analysis of various aspects of animal data
##What is the total number of data elements for each sex? 
```{r}
res1 <- dataset1 %>%  count(sex)
#For debugging res1
ggplot(res1,
       aes(x=res1$n,
           y=res1$sex, fill = res1$sex
           ))+
    geom_bar(stat="identity", position="dodge")
```
##What is the Alphabetical order of all species? 
```{r}
dataset1 %>%  count(sex, species) %>%  arrange(species, desc(n)) %>%  head(10) %>% kable() %>%  kable_styling()
```
##What is the min, mean and max for all species?
```{r}
dataset1 %>% group_by(species_id) %>%
    summarize(
        mean_hindfoot_length = mean(hindfoot_length),
        min_hindfoot_length = min(hindfoot_length),
        max_hindfoot_length = max(hindfoot_length),
        n = n()
    ) %>%  head(10) %>% kable() %>%  kable_styling()
```
##What is the heaviest animal per year?
```{r}
dataset1 %>%    group_by(year) %>%    filter(weight == max(weight)) %>%    select(year, genus, species, weight) %>%    arrange(year)  %>%  head(10) %>% kable() %>%  kable_styling()
```

##What is the animal weight in killograms and pounds?
```{r}
dataset1 %>% mutate(weight_kg = weight / 1000, weight_lb = weight_kg * 2.2) %>%  head(10) %>% kable() %>%  kable_styling()

```

##What is the average weight by sex and species id?
```{r}
res2 <- dataset1 %>% group_by(sex, species_id) %>%  summarize(mean_weight = mean(weight, na.rm = TRUE))

ggplot(res2,
       aes(x=res2$species_id,
           y=res2$mean_weight, fill = res2$sex
           ))+
    geom_bar(stat="identity", position="dodge")
```

##What is the sorted list of data by mean_weight in descending order?
```{r}
dataset1 %>%  group_by(sex, species_id) %>%  summarize(mean_weight = mean(weight), min_weight = min(weight)) %>%  arrange(desc(mean_weight))   %>%  head(10) %>% kable() %>%  kable_styling()
```

##What is mean_weight by genus? 
```{r}
dataset1_01 <- dataset1 %>% group_by(genus, plot_id) %>% summarize(mean_weight = mean(weight)) 
dataset1_01 %>%  head(10) %>% kable() %>%  kable_styling()

```
##What is the mean_weight by plot_id and genus?
```{r}
dataset1_01_spread <- dataset1_01 %>%  spread(key = genus, value = mean_weight, fill=0)
dataset1_01_spread %>%  head(10) %>% kable() %>%  kable_styling()

```

##What is mean_weight by plot_id?

```{r}
dataset1_02_gather <- dataset1_01_spread %>% gather(key = genus, value = mean_weight, -plot_id) 
dataset1_02_gather %>% head(10) %>% kable() %>%  kable_styling()
```
#Dataset 2 -- Steel 
Reference: https://www.census.gov
            
           https://www.census.gov/foreign-trade/Press-Release/2018pr/03/steel/steel2bf.xls
           
#Retrieve Data File for Dataset 2.

```{r}
#https://stackoverflow.com/questions/39110755/skip-specific-rows-using-read-csv-in-r
dataset2 <- read.csv(file="https://raw.githubusercontent.com/AjayArora35/Data-607-Project-2/master/steel2bf (2).csv", header=FALSE, stringsAsFactors = FALSE, skip=15)
head(dataset2,50)

```           
#Data Clean

##Get the first 47 observations X 13 variables
```{r}
#https://stats.idre.ucla.edu/r/modules/subsetting-data/
dataset2 <- dataset2[1:47,1:13]

head(dataset2,100)
```
##Get rid of (-) present in the data; replace with NA
```{r}
dataset2[dataset2 =='(-)'] <- NA
dataset2
```
##Give names to columns
```{r}
names(dataset2) <- c('Countries','032018_quantity','032018_value','022018_quantity','022018_value',
                              '032017_quantity','032017_value', '022017_quantity','022017_value',
                              '03122018_quantity','03122018_value', '03122017_quantity','03122017_value')
head(dataset2)
```

##Get rid of all the commas in the numeric values and convert to numeric data type
```{r}
dataset2$`032018_quantity` <- as.numeric(gsub(",","",dataset2$`032018_quantity`))
dataset2$`032018_value` <- as.numeric(gsub(",","",dataset2$`032018_value`))
dataset2$`022018_quantity` <- as.numeric(gsub(",","",dataset2$`022018_quantity`))
dataset2
```